\documentclass[11pt]{article}

\usepackage{sectsty}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{upquote}
\usepackage{longtable}
\newcommand{\tab}{\hspace*{2em}}

\begin{document}

% title stuff
\title{DotPar \\ Final Report \\ Team 3}
\author{
Logan Donovan - lrd2127@columbia.edu \and
Justin Hines - jph2149@columbia.edu \and
Andrew Hitti - aah2147@columbia.edu \and
Nathan Hwang - nyh2105@columbia.edu \and
Sid Nair - ssn2114@columbia.edu}
\maketitle

%----------------------------
\section{Introduction}
Whole Team
Updated Whitepaper :)

\section{Language Tutorial}
Whole team
\begin{itemize}
\item Make sure all of the programs run
\end{itemize}


\section{Language Reference Manual}
Whole Team
\begin{itemize}
\item Includes complete grammar
\end{itemize}

\section{Project Plan}
Project Manager - Logan
\begin{itemize}
\item State what process used was used to develop the language and its translator.
\item State the roles and responsibilities of each team member.
\item Include the implementation style sheet used by the team.
\item Show the timeline of what was done and when.
\item Include your project log.
\end{itemize}

\section{Language evolution}
Language Guru
\begin{itemize}
\item Describe how the language evolved during the implementation and what steps were used to try to maintain the good attributes of the original language proposal.
\item Describe the compiler tools used to create the compiler components.
\item Describe what unusual libraries are used in the compiler.
\item Describe what steps were taken to keep the LRM and the compiler consistent.
\end{itemize}

\section{Translator architecture}
System Architect
\begin{itemize}
\item Show the architectural block diagram of translator.
\item Describe the interfaces between the modules.
\item State which modules were written by which team members.
\end{itemize}

\section{Development and run-time environment}
System Integrator
\begin{itemize}
\item Describe the software development environment used to create the compiler.
\item Show the makefile used to create and test the compiler during development.
\item Describe the run-time environment for the compiler.
\end{itemize}

\section{Test Plan}
System Tester
\begin{itemize}
\item Describe the test methodology used during the development.
\item Show programs used to test your translator.
\end{itemize}

\section{Conclusions}
Whole team
\begin{itemize}
\item lessons learned as a team
\item lessons for future teams
\item suggestions for the instructor on what topics to keep, drop or add in future courses
\end{itemize}
Individually
\begin{itemize}
\item Lessons learned individually
\end{itemize}

\section{Appendix}
Include a listing of the complete source code with identification of who wrote which module of the compiler. This listing does not have to be included in the paper copy of the final report.


\end{document}